Attribution of undesirable character traits, rather than trait-based dehumanization, predicts punishment decisions

### 2023-06-27 manuscript.Rmd
Replaced ANOVA with t-test in Exp 3a

aov_3a1 <- aov_ez('CaseID',
                 'values',
                 Study3a_mod2,
                 within = c('human'),
                 anova_table = list('pes'))

nice_3a1 <- nice(aov_3a1, 
                es = "pes", 
                correction = "none",
                sig_symbols = rep("",4))

@@

aov_3a2 <- aov_ez('CaseID',
                 'values',
                 Study3a_mod3,
                 within = c('human'),
                 anova_table = list('pes'))

nice_3a2 <- nice(aov_3a2, 
                es = "pes", 
                correction = "none",
                sig_symbols = rep("",4))

@@

A $2 \times 2$ within-subjects ANOVA was conducted to examine how variations in the desirability (desirable or undesirable) and humanness (uniquely human or shared with other animals) of the traits used to describe criminals influenced participants' agreement with granting them parole. In line with our prediction, a significant main effect of trait desirability was found, *F*(`r nice_3a[1,2]`) = `r nice_3a[1,4]`, *p* = `r format.pval(nice_3a[1,6], digits = 3, eps = 0.001, nsmall = 3)`, $\eta_{p}^{2}$ = `r nice_3a[1,5]`. Criminals described with socially undesirable traits (*M* = `r round(undesirable_mean, digits = 2)`, *SE* = `r round(undesirable_se, digits = 2)`) were less likely to be granted parole than were those described with desirable traits (*M* = `r round(desirable_mean, digits = 1)`, *SE* = `r round(desirable_se, digits = 2)`); see Figure \@ref(fig:figure3). A main effect of trait humanness was also found, *F*(`r nice_3a[2,2]`) = `r nice_3a[2,4]`, *p* = `r format.pval(nice_3a[2,6], digits = 3, eps = 0.001, nsmall = 3)`, $\eta_{p}^{2}$ = `r nice_3a[2,5]`. Contrary to the predictions of the dual model, however, criminals who were described with uniquely human traits (*M* = `r round(human_mean, digits = 1)`, *SE*= `r round(human_se, digits = 2)`) were less likely to be granted parole than those described with traits shared with other animals (*M* = `r round(nonhuman_mean, digits = 1)`, *SE* = `r round(nonhuman_se, digits = 2)`). A significant interaction effect between trait humanness and trait desirability was also found, *F*(`r nice_3a[3,2]`) = `r nice_3a[3,4]`, *p* = `r format.pval(nice_3a[3,6], digits = 3, eps = 0.001, nsmall = 3)`, $\eta_{p}^{2}$ = `r nice_3a[3,5]`. Criminals described using undesirable uniquely human traits (*M* = `r round(UH_undesirable_mean, digits = 1)`, *SE* = `r round(UH_undesirable_se, digits = 2)`) were less likely to be granted parole than were criminals described using desirable uniquely human traits (*M* = `r round(UH_desirable_mean, digits = 1)`, *SE* = `r round(UH_desirable_se, digits = 2)`), *F*(`r nice_3a1[1,2]`) = `r nice_3a1[1,4]`, *p* = `r format.pval(nice_3a1[1,6], digits = 3, eps = 0.001, nsmall = 3)`, $\eta_{p}^{2}$ = `r nice_3a1[1,5]`. Similarly, criminals described using undesirable traits shared with other species (*M* = `r round(nonUH_undesirable_mean, digits = 1)`, *SE*= `r round(nonUH_undesirable_se, digits = 2)`) less likely to be granted parole than were those described using desirable traits shared with other species (*M* = `r round(nonUH_desirable_mean, digits = 1)`, *SE* = `r round(nonUH_desirable_se, digits = 2)`), *F*(`r nice_3a2[1,2]`) = `r nice_3a2[1,4]`, *p* = `r format.pval(nice_3a2[1,6], digits = 3, eps = 0.001, nsmall = 3)`, $\eta_{p}^{2}$ = `r nice_3a2[1,5]`.

>>>>

# # replaced ANOVA with t-test as only has two levels
# aov_3a1 <- aov_ez('CaseID',
#                  'values',
#                  Study3a_mod2,
#                  within = c('human'),
#                  anova_table = list('pes'))
# nice_3a1 <- nice(aov_3a1, 
#                 es = "pes", 
#                 correction = "none",
#                 sig_symbols = rep("",4))

hdata <- subset(Study3a_mod2,Study3a_mod2$human=='Human')
nhdata <- subset(Study3a_mod2,Study3a_mod2$human=='Non-Human')
ttest_3a1 <- t.test(hdata$values,nhdata$values,paired=TRUE)
d_3a1 <- mean(hdata$values-nhdata$values)/sd(hdata$values-nhdata$values)

@@

# # replaced ANOVA with t-test as only has two levels
# aov_3a2 <- aov_ez('CaseID',
#                  'values',
#                  Study3a_mod3,
#                  within = c('human'),
#                  anova_table = list('pes'))
# 
# nice_3a2 <- nice(aov_3a2, 
#                 es = "pes", 
#                 correction = "none",
#                 sig_symbols = rep("",4))

hdata <- subset(Study3a_mod3,Study3a_mod3$human=='Human')
nhdata <- subset(Study3a_mod3,Study3a_mod3$human=='Non-Human')
ttest_3a2 <- t.test(hdata$values,nhdata$values,paired=TRUE)
d_3a2 <- mean(hdata$values-nhdata$values)/sd(hdata$values-nhdata$values)

@@

A $2 \times 2$ within-subjects ANOVA was conducted to examine how variations in the desirability (desirable or undesirable) and humanness (uniquely human or shared with other animals) of the traits used to describe criminals influenced participants' agreement with granting them parole. In line with our prediction, a significant main effect of trait desirability was found, *F*(`r nice_3a[1,2]`) = `r nice_3a[1,4]`, *p* = `r format.pval(nice_3a[1,6], digits = 3, eps = 0.001, nsmall = 3)`, $\eta_{p}^{2}$ = `r nice_3a[1,5]`. Criminals described with socially undesirable traits (*M* = `r round(undesirable_mean, digits = 2)`, *SE* = `r round(undesirable_se, digits = 2)`) were less likely to be granted parole than were those described with desirable traits (*M* = `r round(desirable_mean, digits = 1)`, *SE* = `r round(desirable_se, digits = 2)`); see Figure \@ref(fig:figure3). A main effect of trait humanness was also found, *F*(`r nice_3a[2,2]`) = `r nice_3a[2,4]`, *p* = `r format.pval(nice_3a[2,6], digits = 3, eps = 0.001, nsmall = 3)`, $\eta_{p}^{2}$ = `r nice_3a[2,5]`. Contrary to the predictions of the dual model, however, criminals who were described with uniquely human traits (*M* = `r round(human_mean, digits = 1)`, *SE*= `r round(human_se, digits = 2)`) were less likely to be granted parole than those described with traits shared with other animals (*M* = `r round(nonhuman_mean, digits = 1)`, *SE* = `r round(nonhuman_se, digits = 2)`). A significant interaction effect between trait humanness and trait desirability was also found, *F*(`r nice_3a[3,2]`) = `r nice_3a[3,4]`, *p* = `r format.pval(nice_3a[3,6], digits = 3, eps = 0.001, nsmall = 3)`, $\eta_{p}^{2}$ = `r nice_3a[3,5]`. Criminals described using undesirable uniquely human traits (*M* = `r round(UH_undesirable_mean, digits = 1)`, *SE* = `r round(UH_undesirable_se, digits = 2)`) were less likely to be granted parole than were criminals described using desirable uniquely human traits (*M* = `r round(UH_desirable_mean, digits = 1)`, *SE* = `r round(UH_desirable_se, digits = 2)`), *t*(`r ttest_3a1$parameter`) = `r round(ttest_3a1$statistic,digits=2)`, *p* = `r format.pval(ttest_3a1$p.value, digits = 3, eps = 0.001, nsmall = 3)`, $d$ = `r round(d_3a1,digits=2)`. Similarly, criminals described using undesirable traits shared with other species (*M* = `r round(nonUH_undesirable_mean, digits = 1)`, *SE*= `r round(nonUH_undesirable_se, digits = 2)`) less likely to be granted parole than were those described using desirable traits shared with other species (*M* = `r round(nonUH_desirable_mean, digits = 1)`, *SE* = `r round(nonUH_desirable_se, digits = 2)`), *t*(`r ttest_3a2$parameter`) = `r round(ttest_3a2$statistic,digits=2)`, *p* = `r format.pval(ttest_3a2$p.value, digits = 3, eps = 0.001, nsmall = 3)`, $d$ = `r round(d_3a2,digits=2)`.

### 2023-06-27 manuscript.tex
Replaced ANOVA with t-test in Exp 3a

One way to de-confound evidence for trait-based dehumanization from evidence of negative evaluation is to ask participants to rate the target group on traits that are uniquely human but socially undesirable (\protect\hyperlink{ref-Over2020a}{Over, 2020a}, \protect\hyperlink{ref-Over2020b}{2020b}). Previous research has established that undesirable character traits such as jealousy, spite and bitterness are considered unique to humans and socially undesirable (\protect\hyperlink{ref-Enock2021a}{Enock, Tipper, et al., 2021}). Enock and colleagues (\protect\hyperlink{ref-Enock2021a}{2021}) found that whereas participants tend to attribute socially desirable human traits more strongly to the ingroup, they tend to attribute socially undesirable traits more strongly to the outgroup. This result held across three intergroup contexts (see also \protect\hyperlink{ref-Enock2021b}{Enock, Flavell, Tipper, \& Over, 2021}; \protect\hyperlink{ref-Enock2022}{Enock \& Over, 2022}). Enock and colleagues conclude that apparent evidence for trait-based dehumanization may be better explained by intergroup preference. However, it is not yet clear how the attribution of uniquely human character traits relates to harm.

We revisit the hypothesised causal relationship between trait-based dehumanization and harm in the context of endorsing harsh punishment for criminals. In Studies 1A and 1B, we seek to conceptually replicate the key findings of Bastian and colleagues (\protect\hyperlink{ref-Bastian2013}{2013}) suggesting that the extent to which participants animalistically (Study 1A) and mechanistically (Study 1B) dehumanize criminals predicts the severity of the punishment participants endorse for them. In Studies 2A and 2B, we adopt a similar design but incorporate socially undesirable traits into our stimulus set. This addition to the design allows us to investigate whether trait-based dehumanization predicts the severity of punishment, undesirable trait attribution predicts the severity of punishment, or both predict the severity of punishment. In Studies 3A and 3B, we seek to investigate a similar question using an experimental design. We present participants with vignettes in which criminals are described using character traits that differ in how socially desirable they are and in whether or not they are unique to humans and measure how these varying descriptions influence participants' parole decisions. This allows us to directly measure whether there is a causal relationship between trait-based dehumanization and punishment.

@@

All studies received ethical approval from the Psychology Departmental Ethics Committee University of York. All data collection occurred online, and the studies were created and administered using Qualtrics (\url{https://www.qualtrics.com}). Participants were recruited through the online platform Prolific (\url{https://www.prolific.co}), with an independent sample recruited for each study. Informed consent was obtained at the start of each session according to approved ethical guidelines. Inclusion criteria for each study included adult participants fluent in English who had never been to prison for committing a crime and had a Prolific approval rating of at least 90\% (95\% for Studies 3A and 3B). Increases in Prolific's recommended rate of compensation for participation during data collection meant the reward ranged from approximately £7 per hour in Studies 1A and 1B to approximately £8 in the other four studies. Assumption testing and analyses were conducted using \emph{SPSS} and \emph{RStudio}. All studies were pre-registered on AsPredicted.com before commencing data collection. Links to pre-registration documents, data files, files for computational replication on GitHub, and supplementary materials including the stimuli used for each study can be found at: \url{https://osf.io/d4cvp/?view_only=ccf139206e59471b979f53708595ef05}.

@@

A power analysis using G*Power indicated that a sample size of 135 would allow us to detect a medium effect size (\(\eta_p^2\) = .09) with a \(2 \times 2\) factorial design, an alpha of .05 and a power of .95. To counterbalance the sample equally, a sample of 136 participants was collected, of whom 78 identified as female, 55 as male, 2 as non-binary, and 1 who preferred not to indicate their gender identity. Ages ranged from 18 to 63 (\emph{M} = 24.9, \emph{SD} = 6.97). All participants were adults fluent in English who had never been to prison for committing a crime. Due to a noticeable increase in failed attention checks during pilot data collection, the minimal approval rating on Prolific was raised from 90\% to 95\%. Despite this, data from 46 participants were omitted and replaced due to failed attention checks. Three participants were mistakenly recruited after the intended sample size had been met, and thus, their data were excluded from analyses\footnote{Including data submitted by excess participants in analyses yielded the same results as those reported.}. Participation took an average of seven minutes.

@@

\textbf{Attention check}. An additional paragraph was included, largely similar to the other four paragraphs, describing a criminal named Charlie. However, in the middle of the paragraph, the following sentence was included: `\emph{This paragraph is an attention check: please move the slider all the way to Strongly Disagree on the left-hand side}'. Data submitted by any participants who did not respond within 20 points of the instructed end of the 100-point scale were omitted and replaced.

@@

A \(2 \times 2\) within-subjects ANOVA was conducted to examine how variations in the desirability (desirable or undesirable) and humanness (uniquely human or shared with other animals) of the traits used to describe criminals influenced participants' agreement with granting them parole. In line with our prediction, a significant main effect of trait desirability was found, \emph{F}(1, 135) = 369.43, \emph{p} = \textless0.001, \(\eta_{p}^{2}\) = .732. Criminals described with socially undesirable traits (\emph{M} = 38.82, \emph{SE} = 1.72) were less likely to be granted parole than were those described with desirable traits (\emph{M} = 77.8, \emph{SE} = 1.44); see Figure \ref{fig:figure3}. A main effect of trait humanness was also found, \emph{F}(1, 135) = 51.62, \emph{p} = \textless0.001, \(\eta_{p}^{2}\) = .277. Contrary to the predictions of the dual model, however, criminals who were described with uniquely human traits (\emph{M} = 53, \emph{SE}= 1.36) were less likely to be granted parole than those described with traits shared with other animals (\emph{M} = 63.7, \emph{SE} = 1.5). A significant interaction effect between trait humanness and trait desirability was also found, \emph{F}(1, 135) = 54.67, \emph{p} = \textless0.001, \(\eta_{p}^{2}\) = .288. Criminals described using undesirable uniquely human traits (\emph{M} = 27.9, \emph{SE} = 1.93) were less likely to be granted parole than were criminals described using desirable uniquely human traits (\emph{M} = 78, \emph{SE} = 1.7), \emph{F}(1, 135) = 430.39, \emph{p} = \textless0.001, \(\eta_{p}^{2}\) = .761. Similarly, criminals described using undesirable traits shared with other species (\emph{M} = 49.8, \emph{SE}= 2.32) less likely to be granted parole than were those described using desirable traits shared with other species (\emph{M} = 77.6, \emph{SE} = 1.6), \emph{F}(1, 135) = 111.62, \emph{p} = \textless0.001, \(\eta_{p}^{2}\) = .453.

@@

>>>>

One way to de-confound evidence for trait-based dehumanization from evidence of negative evaluation is to ask participants to rate the target group on traits that are uniquely human but socially undesirable (\protect\hyperlink{ref-Over2020a}{Over, 2020a}, \protect\hyperlink{ref-Over2020b}{2020b}). Previous research has established that undesirable character traits such as jealousy, spite and bitterness are considered unique to humans and socially undesirable (\protect\hyperlink{ref-Enock2021a}{Enock, Tipper, et al., 2021}). Enock and colleagues (\protect\hyperlink{ref-Enock2021a}{2021}) found that participants tend to attribute socially desirable human traits more strongly to the ingroup, and socially undesirable traits more strongly to the outgroup. This result held across three intergroup contexts (see also \protect\hyperlink{ref-Enock2021b}{Enock, Flavell, Tipper, \& Over, 2021}; \protect\hyperlink{ref-Enock2022}{Enock \& Over, 2022}). Enock and colleagues conclude that apparent evidence for trait-based dehumanization may be better explained by intergroup preference. However, it is not yet clear how the attribution of uniquely human character traits relates to harm.

We revisit the hypothesised causal relationship between trait-based dehumanization and harm in the context of endorsing harsh punishment for criminals. In Studies 1A and 1B, we seek to conceptually replicate the key findings of Bastian and colleagues (\protect\hyperlink{ref-Bastian2013}{2013}) suggesting that the extent to which participants animalistically (Study 1A) and mechanistically (Study 1B) dehumanize criminals predicts the severity of the punishment participants endorse for them. In Studies 2A and 2B, we adopt a similar design but incorporate socially undesirable traits into our stimulus set. This addition to the design allows us to investigate whether trait-based dehumanization predicts the severity of punishment, undesirable trait attribution predicts the severity of punishment, or both predict the severity of punishment. In Studies 3A and 3B, we seek to investigate a similar question using an experimental design. We present participants with vignettes in which criminals are described using character traits that differ in how socially desirable they are and whether or not they are unique to humans. We also measure how these varying descriptions influence participants' parole decisions. This allows us to directly measure whether there is a causal relationship between trait-based dehumanization and punishment.

@@

All studies received ethical approval from the Psychology Departmental Ethics Committee University of York. All data collection occurred online, and the studies were created and administered using Qualtrics (\url{https://www.qualtrics.com}). Participants were recruited through the online platform Prolific (\url{https://www.prolific.co}), with an independent sample recruited for each study. Informed consent was obtained at the start of each session according to approved ethical guidelines. Inclusion criteria for each study included adult participants fluent in English who had never been to prison for committing a crime and had a Prolific approval rating of at least 90\% (95\% for Studies 3A and 3B). Increases in Prolific's recommended rate of compensation for participation during data collection meant the reward ranged from approximately £7 per hour in Studies 1A and 1B to approximately £8 in the other four studies. Assumption testing and analyses were conducted using \emph{SPSS} and \emph{RStudio}. All studies were pre-registered on AsPredicted.com before commencing data collection. Links to pre-registration documents, data files (\protect\hyperlink{ref-Brennan2023}{Brennan, Enock, \& Over, 2023}), a fully computationally reproducible version of the manuscript, and supplementary materials including the stimuli used for each study can be found at: \url{https://osf.io/d4cvp/?view_only=ccf139206e59471b979f53708595ef05}.

@@

A power analysis using G*Power, with effect size specification as in SPSS, indicated that a sample size of 135 would allow us to detect a medium effect size (\(\eta_p^2\) = .09) with a \(2 \times 2\) factorial, repeated measures design, an alpha of .05, and a power of .95. To counterbalance the sample equally, a sample of 136 participants was collected, of whom 78 identified as female, 55 as male, 2 as non-binary, and 1 who preferred not to indicate their gender identity. Ages ranged from 18 to 63 (\emph{M} = 24.9, \emph{SD} = 6.97). All participants were adults fluent in English who had never been to prison for committing a crime. Due to a noticeable increase in failed attention checks during pilot data collection, the minimal approval rating on Prolific was raised from 90\% to 95\%. Despite this, data from 46 participants were omitted and replaced due to failed attention checks. Three participants were mistakenly recruited after the intended sample size had been met, and thus, their data were excluded from analyses\footnote{Including data submitted by excess participants in analyses yielded the same results as those reported.}. Participation took an average of seven minutes.

@@

\textbf{Attention check}. An additional paragraph describing a criminal named Charlie was included, largely similar to the other four paragraphs. However, in the middle of the paragraph, the following sentence was included: `\emph{This paragraph is an attention check: please move the slider all the way to Strongly Disagree on the left-hand side}'. Data submitted by any participants who did not respond within 20 points of the instructed end of the 100-point scale were omitted and replaced.

@@

A \(2 \times 2\) within-subjects ANOVA was conducted to examine how variations in the desirability (desirable or undesirable) and humanness (uniquely human or shared with other animals) of the traits used to describe criminals influenced participants' agreement with granting them parole. In line with our prediction, a significant main effect of trait desirability was found, \emph{F}(1, 135) = 369.43, \emph{p} = \textless0.001, \(\eta_{p}^{2}\) = .732. Criminals described with socially undesirable traits (\emph{M} = 38.82, \emph{SE} = 1.72) were less likely to be granted parole than were those described with desirable traits (\emph{M} = 77.8, \emph{SE} = 1.44); see Figure \ref{fig:figure3}. A main effect of trait humanness was also found, \emph{F}(1, 135) = 51.62, \emph{p} = \textless0.001, \(\eta_{p}^{2}\) = .277. Contrary to the predictions of the dual model, however, criminals who were described with uniquely human traits (\emph{M} = 53, \emph{SE}= 1.36) were less likely to be granted parole than those described with traits shared with other animals (\emph{M} = 63.7, \emph{SE} = 1.5). A significant interaction effect between trait humanness and trait desirability was also found, \emph{F}(1, 135) = 54.67, \emph{p} = \textless0.001, \(\eta_{p}^{2}\) = .288. Criminals described using undesirable uniquely human traits (\emph{M} = 27.9, \emph{SE} = 1.93) were less likely to be granted parole than were criminals described using desirable uniquely human traits (\emph{M} = 78, \emph{SE} = 1.7), \emph{t}(135) = 20.75, \emph{p} = \textless0.001, \(d\) = 1.78. Similarly, criminals described using undesirable traits shared with other species (\emph{M} = 49.8, \emph{SE}= 2.32) less likely to be granted parole than were those described using desirable traits shared with other species (\emph{M} = 77.6, \emph{SE} = 1.6), \emph{t}(135) = 10.57, \emph{p} = \textless0.001, \(d\) = 0.91.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Brennan2023}{}}%
Brennan, R. A., Enock, F., \& Over, H. (2023). \emph{Datasets for "attribution of undesirable character traits, rather than trait-based dehumanization, predicts punishment decisions"} {[}{[}Datasets{]}{]}. Retrieved from \url{https://osf.io/d4cvp/?view_only=ccf139206e59471b979f53708595ef05}
